Contextual recruitment of selective attention can be updated via changes in task-relevance

### 2019-10-17 bWFudXNjcmlwdF9wcmV0dHkuUm1k.Rmd
draft complete

abstract: "Evidence across a wide variety of attention paradigms shows that environmental cues can trigger adjustments to ongoing priorities for attending to relevant and irrelevant information. This context-specific control over attention suggests that cognitive control can be both automatic and flexible. For instance, in selective attention tasks, congruency effects are larger for items that appear in a context associated with infrequent conflict than in a context associated with frequent conflict. Since the to-be-presented context cannot be predicted or prepared for in advance, attention is assumed to be rapidly updated on-the-fly, triggered by the currently presented context. Context-specific control exemplifies how learning and memory processes can influence attention to enable cognitive flexibility. However, what determines the use of previously learned associations still remains unclear. In the current study, we examined whether task-relevance would influence the learning and use of context cues in a flanker task. Using a secondary counting task, context dimensions associated with differing levels of conflict were made task-relevant or -irrelevant across the experiment. In short, we found that making new contextual information task-relevant caused participants to ignore a previously learned context-attention association and adopt a new context-specific control strategy; all without changing the experimental stimuli. Furthermore, we found participants did not spontaneously learn about context-specific proportion manipulations and explicit instructions insufficient for producing context-specific effects. These results suggests that task-relevance is a key determinant of context-specific control."

@@

bibliography      : ["task-relevance-cspc.bib"]

@@

header-includes:
  - \usepackage{setspace}\singlespacing 
  # - \newenvironment{zeroindent}
  #    {\par\setlength{\parindent}{5pt}}
  #    {\par}  
  - \setlength{\parskip}{0pt}
  - \raggedbottom

@@

For example, Crump, Gong, and Milliken [@crump_context-specific_2006; see also, @corballis_independent_2003] presented Stroop stimuli in one of two randomly chosen locations and manipulated the frequency of conflict associated with each location. One location was associated with a high frequency of conflict (25% congruent trials) and the other with a low frequency of conflict (75% congruent trials). Overall, the proportion of congruent trials was 50% and randomized such that the upcoming location could not be predicted. Even so, congruency effects were shown to be smaller for trials where the stimulus appeared in the high conflict location as compared to the low conflict location. This effect, known now as the context-specific proportion congruent effect (CSPC), has now been replicated in a number of different selective attention paradigms [e.g., @alards-tomalin_auditory_2017; @blais_trial-by-trial_2015; @bugg_conflict-triggered_2014; @crump_context-dependent_2018; @crump_learning_2016; @fischer_context-sensitive_2014; @hubner_location-specific_2016].

@@

Real-world environments are complex and contain many potential cues for directing attention. For instance, a coffee shop is a rich environment with numerous context cues (e.g., pictures on the wall, chairs, coffee, laptop), each of which someone could have different experiences with and afford different learned associations. A laptop for example, could cue attentional processing helpful for writing while other features of the coffee shop might cue attentional processing related to other experiences like socializing and people-watching. Given that all of these cues are available to set attentional priorities, what determines which associative relationships will be used to guide attention? On the one hand, context-based retrieval of attentional priorities might be completely obligatory: the mere presence of learned cues might trigger attention in a ballistic manner [e.g., @schneider_controlled_1977; @shiffrin_controlled_1977]. Attention in this case would be at the mercy of the environment and attention is guided by the strongest association. On the other hand, our current task-goals might constrain the expression of previously learned associations. If someone sits down to write in the coffee shope, their current goals may allow the laptop to take priority over other environmental cues that are irrelevant to the task at hand.

@@

An alternative view---the memory-based perspective---takes a more nuanced view [@norman_memory_1969; @norman_memory_1976; @abrahamse_grounding_2016; @brosowsky_memory-guided_2018; @crump_learning_2016; @egner_creatures_2014]. Here, practice is only important to the extent that it improves memory encoding and retrieval. Learning an association can be thought of as the process of encoding experiences where attentional priorities and cues co-occur. The expression of previously learned associations however, is dictated by memory retrieval---a similarity-based, cue-driven process [e.g., @jacoby_interpreting_1978; @hintzman_minerva_1984; @hintzman_schema_1986; @hintzman_judgments_1988]. Presumably, the features of our immediate environment automatically cue memory and reinstate the retrieved attentional priorites [@logan_toward_1988; @logan_attention_1992]. If a few prior experiences are easily retrievable (i.e., memories are distinct or very similar to the retrieval cue), they could support the reinstatement of a learned attentional association despite very little practice [e.g., @brosowsky_memory-guided_2018]. Conversely, if prior experiences are difficult to retrieve (i.e., memories are indistinct or low in similarity to the retrieval cue), then learned associations may not be expressed, even after much practice. To the extent that you could manipulate memory retrieval, you should als be able manipulate the expression of learned associations.

Here, we examine whether task-relevance influences the retrieval of previously used attentional priorities. We hypothesized that the cue-driven, memory-retrieval process might be constrained to the features of our environment that are relevant to our on-going task and goals. If this is the case, then a previously learned association should be easily ignored if that feature was made irrelevant to the task because that feature would no longer contribute to memory-retrieval. Furthermore, it would suggest that the mere presence of a cue is not sufficient for automatically directing attention, but instead requires task-integration.

@@

Prior work suggests that the task-relevance of contextual information could be important for establishing new associations between context cues and attentional priorities. Simply put, CSPC effects tend to emerge when context cues are somehow made relevant to the task, but do not when made irrelevant [@canadas_social_2013; @crump_context-specific_2006; @crump_context-specific_2008]. However, what determines the  expression of previously learned context-attention associations remains unclear. In the current study, we examined whether manipulating task-relevance would allow participants to ignore a previously learned association and adopt a new context-specific control strategy.

@@

The critical measure of interest is the congruency effect produced in the frequency unbiased context. If, on the one hand, task-relevance has no impact on the use of context-attention associations we would expect no differences between the congruency effects when the high conflict context dimension is made task-relevant or the low conflict context dimension is made task-relevant. This could occur because all three contexts are treated as individual contexts throughout the whole experiment [e.g., @canadas_social_2013]. Or it could occur because associations formed in the first block interfere with learning new associations in the second [e.g., @brosowsky_context-specific_2016]. On other hand, task-relevance may dictate the expression of previously learned associations. In this case, we would expect smaller congruency effects when the high conflict context dimension is made task-relevant as compared to when the low conflict context dimension is made task-relevant. This would demonstrate that participants were able to ignore a previously learned association and adopt a new context-specific control strategy.

@@

We also conduncted an additional, more severe, test of the list-shift hypothesis. Given that the asymmetry is interpreted as resulting from a carry-over of learned attentional modulation across phases, we might expect that any asymmetry would be exaggerated near the point of transition. We therefore repeated the above analysis but restricted it to the 24 trials immediately before the transition and 24 trials immediately after the transition.

@@

In Experiment 1, we found that manipulating the task-relevance of a shared contextual feature was associated in predictable changes in context-specific proportion congruency effects for frequency unbiased context items. Namely, we found larger congruency effects when the context was associated with low conflict than when the context was associated with high conflict. This suggests that participants learned the association between context and conflict-level in the first block (despite multiple conetxt cues), and subsequently ignored that association in the second block to learn a new association. Task-relevance then, may not just be important for establishing new associations [e.g.,@crump_flexibility_2009] but also important for determining which learned associations are expressed in the present moment.

@@

Second, we assumed that the biased proportion congruency manipulations would have no systematic effect on the unbiased items without the counting task. Although we randomized all aspects of the task across participants to control for potential confounds, it still remains unclear whether the biased items would influence how paricipants responded to the unbiased items.

@@

The goal of the current study has been to examine the importance of task-relevance for the contextual recruitment of selective attention. In Experiment 1, we found that task-relevance allowed participants to exploit context cues in the presence of a competing, contradictory cue and subsequently ignore a previously learned association in favor of a different context cue. In Experiment 2, we conceptually replicated previous work showing that task-relevance was necessary for participants to learn about the biased proportion congruency associations [e.g., @crump_flexibility_2009] and ruled out possible confounding effects of the biased items.

@@

# General Discussion
The aim of the current study was to determine whether task-relevance plays a role in the contextual recruitment of selective attention. In Experiment 1, we tested whether manipulating the relative task-relevance of context cues could cause participants to ignore a previously learned context-association and apply a new association. We created a frequency unbiased context cue that shared features with two frequency biased contexts and used a feature-counting task to manipulate the task-relevance of context dimensions across two blocks of trials. Critically, halfway through the experiment participants received new instructions changing the task-relevant feature from one frequency biased cue to the other.

The key finding was that congruency effects for the frequency unbiased items were significantly larger when the low conflict context was made task-relevant as compared to when the high conflict context was made task-relevant. This result is consistent with prior CSPC effects and, like the previous work, suggests that the context cues triggered rapid adjustments to attentional control [@crump_flexibility_2009]. However, unlike prior studies, we were able to experimentally manipulate the CSPC effect across blocks of trials without changing any of the physical properties of the stimuli. This novel finding demonstrates that participants were able to learn and apply one context-attention association in the first phase, and subsequently ignore that association to learn a new association in the second phase.

This result implicates an important role for task-relevance in producing CSPC phenomena. @crump_context-specific_2008 used shapes as context cues in a prime-probe Stroop task and did not find CSPC effects until the context cues were made task-relevant. Similarly, @canadas_social_2013 eliminated the CSPC effect by making the contextual cue effectively unrelated to the task. These studies suggest that task-relevance plays an important role in establishing new associations between contextual information and attentional priorities to produce CSPC effects. Our finding extends this work in two important ways. First, we show that changing the task-relevance of the presented cues corresponded with a change in attentional control in the predicted direction. This demonstrates that task-relevance is also a key determinant of whether a previously learned attention-context association will be used or ignored in favor of a new association. Second, we show that task-relevance allowed participants resolve competition between two competing contextual cues, responding on the basis of one at the expense of the other. To our knowledge, this is the first demonstration that CSPC effects can be produced when there are multiple, overlapping contextual cues available.

@@

In light of prior work, we take these results as evidence that the contextual recruitment of selective attention, although likely implicit, is not obligatory [e.g., @brosowsky_context-specific_2016], requiring that environmental information be incorporated into the task representation. Similarly, there appears to be flexibility in which environmental features are selected and used to guide attention, which can be rapidly updated depending on the task-relevance of those features (for a similar perspective, see @hommel_event_2004; @hommel_theory_2019). Such a result lends some insight into how context-specific control might operate in more complex, real-world environments, where there is an over-abundance of environmental features that afford many different learned associations. From a theoretical perspective, this result is consistent memory-based accounts of CSPC phenomena [@brosowsky_memory-guided_2018; @bugg_converging_2013; @crump_context-dependent_2018; @crump_flexibility_2009; @crump_reproducing_2017]. Under this view, a memory process encodes attentional priorities in the representation of individual experiences and, as a result, become associated with the environment where they were used. The subsequent reoccurrence of a prior context triggers the retrieval and reinstatement of those attentional priorities. Our results show however, that all the features of the environment may not be treated equally and that only task-relevant features are used to probe memory and guide attention.

>>>>

abstract: "Evidence across a wide variety of attention paradigms shows that environmental cues can trigger adjustments to ongoing priorities for attending to relevant and irrelevant information. This context-specific control over attention suggests that cognitive control can be both automatic and flexible. For instance, in selective attention tasks, congruency effects are larger for items that appear in a context associated with infrequent conflict than in a context associated with frequent conflict. Since the to-be-presented context cannot be predicted or prepared for in advance, attention is assumed to be rapidly updated on-the-fly, triggered by the currently presented context. Context-specific control exemplifies how learning and memory processes can influence attention to enable cognitive flexibility. However, what determines the use of previously learned associations still remains unclear. In the current study, we examined whether task-relevance would influence the learning and use of context cues in a flanker task. Using a secondary counting task, context dimensions associated with differing levels of conflict were made task-relevant or -irrelevant across the experiment. In short, we found that making new contextual information task-relevant caused participants to suppress a previously learned context-attention association and adopt a new context-specific control strategy; all without changing the experimental stimuli. Furthermore, we found participants did not spontaneously learn about context-specific proportion manipulations and explicit instructions were insufficient for producing context-specific effects. These results suggests that task-relevance is a key determinant of context-specific control."

@@

bibliography      : ["task-relevance-cspc.bib"]
appendix          : "appendix_a.rmd"

@@

header-includes:
  - \usepackage{threeparttable}
  - \usepackage{setspace}\singlespacing 
  # - \newenvironment{zeroindent}
  #    {\par\setlength{\parindent}{5pt}}
  #    {\par}  
  - \setlength{\parskip}{0pt}
  - \raggedbottom

@@

For example, Crump, Gong, and Milliken [@crump_context-specific_2006; see also, @corballis_independent_2003] presented Stroop stimuli in one of two randomly chosen locations and manipulated the frequency of conflict associated with each location. One location was associated with a high frequency of conflict (25% congruent trials) and the other with a low frequency of conflict (75% congruent trials). Overall, the proportion of congruent trials was 50% and trials were randomized such that the upcoming location could not be predicted. Even so, congruency effects were shown to be smaller for trials where the stimulus appeared in the high conflict location as compared to the low conflict location. This effect, known now as the context-specific proportion congruent effect (CSPC), has now been replicated in a number of different selective attention paradigms [e.g., @alards-tomalin_auditory_2017; @blais_trial-by-trial_2015; @bugg_conflict-triggered_2014; @crump_context-dependent_2018; @crump_learning_2016; @fischer_context-sensitive_2014; @hubner_location-specific_2016].

@@

Real-world environments are complex and contain many potential cues for directing attention. For instance, a coffee shop is a rich environment with numerous context cues (e.g., pictures on the wall, chairs, coffee, laptop), each of which someone could have different experiences with and afford different learned associations. A laptop for example, could cue attentional processing helpful for writing while other features of the coffee shop might cue attentional processing related to other experiences like socializing and people-watching. Given that all of these cues are available to set attentional priorities, what determines which associative relationships will be used to guide attention? On the one hand, context-based retrieval of attentional priorities might be completely obligatory: the mere presence of learned cues might trigger attention in a ballistic manner [e.g., @schneider_controlled_1977; @shiffrin_controlled_1977]. Attention in this case would be at the mercy of the environment and attention is guided by the strongest association. On the other hand, our current task-goals might constrain the expression of previously learned associations. If someone sits down to write in the coffee shop, their current goals may allow the laptop to take priority over other environmental cues that are irrelevant to the task at hand.

@@

An alternative view---the memory-based perspective---takes a more nuanced view [@norman_memory_1969; @norman_memory_1976; @abrahamse_grounding_2016; @brosowsky_memory-guided_2018; @crump_learning_2016; @egner_creatures_2014]. Here, practice is only important to the extent that it improves memory encoding and retrieval. Learning an association can be thought of as the process of encoding experiences where attentional priorities and cues co-occur. The expression of previously learned associations however, is dictated by memory retrieval---a similarity-based, cue-driven process [e.g., @jacoby_interpreting_1978; @hintzman_minerva_1984; @hintzman_schema_1986; @hintzman_judgments_1988]. Presumably, the features of our immediate environment automatically cue memory and reinstate the retrieved attentional priorities [@logan_toward_1988; @logan_attention_1992]. If a few prior experiences are easily retrievable (i.e., memories are distinct or very similar to the retrieval cue), they could support the reinstatement of a learned attentional association despite very little practice [e.g., @brosowsky_memory-guided_2018]. Conversely, if prior experiences are difficult to retrieve (i.e., memories are indistinct or low in similarity to the retrieval cue), then learned associations may not be expressed, even after much practice. To the extent that you could manipulate memory retrieval, you should also be able manipulate the expression of learned associations.

Here, we examine whether task-relevance influences the retrieval of previously used attentional priorities. We hypothesized that the cue-driven, memory-retrieval process might be constrained to the features of our environment that are relevant to our on-going task and goals. If this is the case, then a previously learned association should be easily suppressed if that feature was made irrelevant to the task because that feature would no longer contribute to memory-retrieval. Furthermore, it would suggest that the mere presence of a cue is not sufficient for automatically directing attention, but instead requires task-integration (for a similar perspective, see @hommel_event_2004; @hommel_theory_2019).

@@

Prior work suggests that the task-relevance of contextual information could be important for establishing new associations between context cues and attentional priorities. Simply put, CSPC effects tend to emerge when context cues are somehow made relevant to the task, but do not when made irrelevant [@canadas_social_2013; @crump_context-specific_2006; @crump_context-specific_2008]. However, what determines the  expression of previously learned context-attention associations remains unclear. In the current study, we examined whether manipulating task-relevance would allow participants to suppress a previously learned association and adopt a new context-specific control strategy.

@@

The critical measure of interest is the congruency effect produced in the frequency unbiased context. If, on the one hand, task-relevance has no impact on the use of context-attention associations we would expect no differences between the congruency effects when the high conflict context dimension is made task-relevant or the low conflict context dimension is made task-relevant. This could occur because all three contexts are treated as individual contexts throughout the whole experiment [e.g., @canadas_social_2013]. Or it could occur because associations formed in the first block interfere with learning new associations in the second [e.g., @brosowsky_context-specific_2016]. On other hand, task-relevance may dictate the expression of previously learned associations. In this case, we would expect smaller congruency effects when the high conflict context dimension is made task-relevant as compared to when the low conflict context dimension is made task-relevant. This would demonstrate that participants were able to suppress a previously learned association and adopt a new context-specific control strategy.

@@

We also conducted an additional, more severe, test of the list-shift hypothesis. Given that the asymmetry is interpreted as resulting from a carry-over of learned attentional modulation across phases, we might expect that any asymmetry would be exaggerated near the point of transition. We therefore repeated the above analysis but restricted it to the 24 trials immediately before the transition and 24 trials immediately after the transition.

@@

In Experiment 1, we found that manipulating the task-relevance of a shared contextual feature was associated in predictable changes in context-specific proportion congruency effects for frequency unbiased context items. Namely, we found larger congruency effects when the context was associated with low conflict than when the context was associated with high conflict. This suggests that participants learned the association between context and conflict-level in the first block (despite multiple context cues), and subsequently suppressed in the second block, allowing participants to learn a new association. Task-relevance then, may not just be important for establishing new associations [e.g.,@crump_flexibility_2009] but also important for determining which learned associations are expressed in the present moment.

@@

Second, we assumed that the biased proportion congruency manipulations would have no systematic effect on the unbiased items without the counting task. Although we randomized all aspects of the task across participants to control for potential confounds, it still remains unclear whether the biased items would influence how participants responded to the unbiased items.

@@

The goal of the current study has been to examine the importance of task-relevance for the contextual recruitment of selective attention. In Experiment 1, we found that task-relevance allowed participants to exploit context cues in the presence of a competing, contradictory cue and subsequently suppress a previously learned association in favor of a different context cue. In Experiment 2, we conceptually replicated previous work showing that task-relevance was necessary for participants to learn about the biased proportion congruency associations [e.g., @crump_flexibility_2009] and ruled out possible confounding effects of the biased items.

@@

# General Discussion
The aim of the current study was to determine whether task-relevance plays a role in the contextual recruitment of selective attention. In Experiment 1, we tested whether manipulating the relative task-relevance of context cues could cause participants to suppress a previously learned context-association and apply a new association. We created a frequency unbiased context cue that shared features with two frequency biased contexts and used a feature-counting task to manipulate the task-relevance of context dimensions across two blocks of trials. Critically, halfway through the experiment participants received new instructions changing the task-relevant feature from one frequency biased cue to the other.

The key finding was that congruency effects for the frequency unbiased items were significantly larger when the low conflict context was made task-relevant as compared to when the high conflict context was made task-relevant. This result is consistent with prior CSPC effects and, like the previous work, suggests that the context cues triggered rapid adjustments to attentional control [@crump_flexibility_2009]. However, unlike prior studies, we were able to experimentally manipulate the CSPC effect across blocks of trials without changing any of the physical properties of the stimuli. This novel finding demonstrates that participants were able to learn and apply one context-attention association in the first phase, and subsequently suppress that association to learn a new association in the second phase.

This result implicates an important role for task-relevance in producing CSPC phenomena. @crump_context-specific_2008 used shapes as context cues in a prime-probe Stroop task and did not find CSPC effects until the context cues were made task-relevant. Similarly, @canadas_social_2013 eliminated the CSPC effect by making the contextual cue effectively unrelated to the task. These studies suggest that task-relevance plays an important role in establishing new associations between contextual information and attentional priorities to produce CSPC effects. Our finding extends this work in two important ways. First, we show that changing the task-relevance of the presented cues corresponded with a change in attentional control in the predicted direction. This demonstrates that task-relevance is also a key determinant the previously learned attention-context association is expressed or suppressed. Second, we show that task-relevance allowed participants resolve competition between two competing contextual cues, responding on the basis of one at the expense of the other. To our knowledge, this is the first demonstration that CSPC effects can be produced when there are multiple, overlapping contextual cues available.

@@

In light of prior work, we take these results as evidence that the contextual recruitment of selective attention, although likely implicit, is not obligatory [e.g., @brosowsky_context-specific_2016], requiring that environmental information be incorporated into the task representation. Similarly, there appears to be flexibility in which environmental features are selected and used to guide attention, which can be rapidly updated depending on the task-relevance of those features (for a similar perspective, see @hommel_event_2004; @hommel_theory_2019). Such a result lends some insight into how context-specific control might operate in more complex, real-world environments, where there is an over-abundance of environmental features that afford many different learned associations. From a theoretical perspective, this result is consistent memory-based accounts of CSPC phenomena [@brosowsky_memory-guided_2018; @bugg_converging_2013; @crump_context-dependent_2018; @crump_flexibility_2009; @crump_reproducing_2017]. Under this view, a memory process encodes attentional priorities in the representation of individual experiences and, as a result, become associated with the environment where they were used. The subsequent re-occurrence of a prior context triggers the retrieval and reinstatement of those attentional priorities. Our results show however, that all the features of the environment may not be treated equally and that only task-relevant features are used to probe memory and guide attention.
